\chapter{Developing \glc}

The following is a quickstart guide to making changes to the \glc\ source code and contributing them back to the project. Note that the preferred method to do this is through \href{https://bitbucket.org}{\normalfont \scshape BitBucket}.

\section{Getting Started}

It's easy to begin working with and changing the \glc\ source code. Assuming you have Mercurial (``{\normalfont \ttfamily hg}'') installed, just do:
\begin{verbatim}
 hg clone https://abensonca@bitbucket.org/abensonca/galacticus galacticus
\end{verbatim}
and you have a cloned copy of the \glc\ repository in the {\normalfont \ttfamily galacticus} directory.

\subsection{Using {\normalfont \scshape BitBucket}}

If you plan to contribute changes back to the \glc] project (please do!), you should consider using \href{https://bitbucket.org}{\normalfont \scshape BitBucket}. After you've created an account for yourself at {\normalfont \scshape BitBucket}, you can ``fork'' the \glc\ repository to have your own working copy. This can be done as follows:
\begin{itemize}
 \item visit the \glc\ repository on {\normalfont \ttfamily BitBucket} at \href{https://bitbucket.org/abensonca/galacticus/overview}{\normalfont \ttfamily https://bitbucket.org/abensonca/galacticus/overview};
 \item click on the ``Fork'' button---you'll be presented with a form;
 \item fill out the form (setting a name for your fork, a short description, etc.), then click the ``Fork repository'' button;
 \item after the fork completes, you'll be taken to the overview page for your forked repository.
\end{itemize}. 

You'll now want to clone this forked repository to your local system. Click on the ``Clone'' button and copy the {\normalfont \ttfamily hg clone} command presented there (you want the {\normalfont \ttfamily SSH} version so that you can push changes back to this repository). Run this command on your local system to get a cloned copy of your new repository. You can now work with this repository, make any changes, and commit then. We'll discuss how to send these changes back to {\normalfont \scshape BitBucket}, and back to the \glc\ project below.

\section{Making Simple Changes}

If you want to make some relatively minor changes to the \glc\ code, such as fixing a typo, adding a new filter, etc. you can just make changes directly on the {\normalfont \ttfamily default} branch (i.e. at the point of active development). To do this, make sure you're at {\normalfont \ttfamily default}:
\begin{verbatim}
 hg pull
 hg update default
\end{verbatim}
Then make your changes, add new files, etc. Once you're done, first check if there have been any changes to {\normalfont \ttfamily default} since you {\normalfont \ttfamily pull}ed:
\begin{verbatim}
 hg incoming
\end{verbatim}
If any new changesets are shown, us {\normalfont \ttfamily hg pull -u} to merge these in to your working copy. Then commit your changes:
\begin{verbatim}
 hg commit
\end{verbatim}
Your changes are now commited to your cloned repository.

\subsection{Contributing Your Changes Back To \glc}

Once you've committed your changes, you can contribute them back to the \glc\ project.

\subsubsection{Via E-mail}

If you just cloned the \glc\ repository directly you can send a patch containing your changes by e-mail to \href{mailto:abenson@carnegiescience.edu}{abenson{@}carnegiescience.edu}. First, create the patch file using:
\begin{verbatim}
 hg export -r begin:end > changes.diff
\end{verbatim}
where {\normalfont \ttfamily begin} and {\normalfont \ttfamily end} are the first and last revisions that you want to include (you can specify more complicated sets of revisions of course). Then simply attached the {\normalfont \ttfamily changes.diff} file to an e-mail. It will be merged into the \glc\ project using
\begin{verbatim}
 hg import changes.diff
\end{verbatim}

\subsubsection{Using {\normalfont \scshape BitBucket}}

If you forked the \glc\ repository on {\normalfont \scshape BitBucket}, you can now push your changes back to {\normalfont \ttfamily BitBucket} using
\begin{verbatim}
 hg push
\end{verbatim}
If you want to contribute these changes back to the \glc\ project the best way to do so is to create a ``pull request''. Simply visit your forked respository on {\normalfont \scshape BitBucket} and click the ``Pull request'' button. The form you're presented with allows you to choose which branch in your respository you want to send changes from, and which branch in the \glc\ project you want them contributed to. Add a title and description of your changes (and, optionally, check the ``Close branch'' box if you're done with this branch of development) then click ``Create pull request''. Assuming your code looks godo and works, it can then be pulled into the \glc\ project.

\section{Making Bigger Changes}

For bigger changes, particularly those where you're adding a new feature, we recommend using Mercurial's ``feature branches''. These provide a permanent record of for which feature each changeset was added. Using feature branches is straightforward. Begin with {\normalfont \ttfamily default} and create a new branch:
\begin{verbatim}
 hg update default
 hg branch myNewFeature
\end{verbatim}
where {\normalfont \ttfamily myNewFeature} is a name for your feature branch. Then begin working, make changes, add new files etc. You can make commits when necessary (and it's good to make several small commits rather than one big one). You should merge {\normalfont \ttfamily default} into your feature branch as often as possible to avoid them getting out of sync (which makes for difficulty later when you want to merge your feature branch back into {\normalfont \ttfamily default}):
\begin{verbatim}
 hg update myFeatureBranch
 hg merge default
 hg commit -m "merged default into myFeatureBranch"
\end{verbatim}
Once the feature branch is stable, you can merge it back into {\normalfont \ttfamily default}:
\begin{verbatim}
 hg update default
 hg merge myFeatureBranch
 hg commit -m "merged myFeatureBranch"
\end{verbatim}
Once you're done developing this feature, you should close the feature branch:
\begin{verbatim}
 hg commit --close-branch -m "finished my feature"
\end{verbatim}
Note that you can always go back and work on a feature branch later, after you have closed it. Just do:
\begin{verbatim}
 hg up myFeatureBranch
 hg merge default
 hg commit -m "merged default into myFeatureBranch"
\end{verbatim}
then continue to work with your feature branch as normal (don't forget to close it again when you're finished working with it).

\section{Releases}

Each release of \glc\ exists as a separate branch within the main \glc\ repository. To work with a particular release use
\begin{verbatim}
 hg update v0.9.2
\end{verbatim}
replacing the version number with whichever version you want. To get back to the development tip use
\begin{verbatim}
 hg update default
\end{verbatim}

\subsection{Bug Fixes In Releases}

To make a bugfix in a release, simply {\normalfont \ttfamily hg update} to that release, fix the bug, and commit your changes. In many cases you'll want to fix the same bug in later releases and also in {\normalfont \ttfamily default}. To do that, just {\normalfont \ttfamily hg update} to each branch in turn, use {\normalfont \ttfamily hg merge fixedBranch} (where ``{\normalfont \ttfamily fixedBranch}'' is the name of the branch in which you fixed the branch, and then commit the merge. Once the bug has been fixed you can contribute the fix back to the \glc\ project using the methods described above.

\section{The \glc\ Build System}

\glc\ use a GNU Make-based build system. Dependencies are automatically discovered, and extensive meta-programming and preprocessing of source files is undertaken as part of the build. As a result, the build system is complicated. In this section the build system is described and detailed.

\subsection{The Makefile}

Files generated during the build (with the exception of final executables) are written to the directory specified by the {\normalfont \ttfamily BUILDPATH} environment variable. This is normally {\normalfont \ttfamily work/build/} (and is set by the {\normalfont \ttfamily Makefile}) unless a special build configuration is requested.

\subsection{Automatic Discovery}

Interdependencies between source files, together with requirements for auto-generated code are automatically discovered during the build process by processing of source files. All such automatic discovery is described below.

\subsubsection{Code Directive Parsing}\label{sec:buildDiscoveryDirectives}

The modular nature of \glc\ is enabled through the use of directives embedded within the source code (XML documents embedded as comment lines starting with a special tag) which instruct the build system how to link together the various functions. Parsing of these directives is handled by the {\normalfont \ttfamily codeDirectivesParse.pl}\index{codeDirectivesParse.pl@{\normalfont \ttfamily codeDirectivesParse.pl}} script.

For {\normalfont \ttfamily include} directives (which generate files that get included into the source), the script generates {\normalfont \ttfamily \$(BUILDPATH)/*.xml} files which describe the directive, and {\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Directives} which contains rules for building those include files.

The script also outputs a file {\normalfont \ttfamily \$(BUILDPATH)/directiveLocations.xml} which provides list of files that contain each directive. This is used by other scripts to permit rapid processing of files associated with each directive.

\subsubsection{Allocatable Arrays}\label{sec:buildDiscoverAllocatables}

Types and ranks of allocatable arrays required by the \glc\ are discovered by the {\normalfont \ttfamily allocatableArrays.pl}\index{allocatableArrays.pl@{\normalfont \ttfamily allocatableArrays.pl}} script. Source files are parsed for dimensionful, allocatable intrinsic types. All such combinations of type and rank are stored to a file {\normalfont \ttfamily \$(BUILDPATH)/allocatableArrays.xml}. The content of this file is later used to build appropriate memory allocation and deallocation functions (see \S\ref{sec:buildGenerationMemory}).

\subsubsection{Executable Files}\index{files!executable}\index{executable files}\label{sec:buildExecutables}

Files which produce executables (including the {\normalfont \ttfamily Galacticus.exe} executable) are discovered by the {\normalfont \ttfamily scripts/build/findExecutables.pl}\index{findExecutables.pl@{\normalfont \ttfamily findExecutables.pl}} script, which generates rules describing how these files should be built, along with all of their dependencies, and writes them to {\normalfont \ttfamily \$(BUILDPATH)/Makefile\_All\_Execs}\index{Makefile_All_Execs@{\normalfont \ttfamily Makefile\_All\_Execs}}. All Fortran files in the {\normalfont \ttfamily source} directory are parsed, and executable files are identified as those containing a {\normalfont \ttfamily program} statement. All executables so identified are also added to the list of objects that will be built by {\normalfont \ttfamily make all}.

\subsubsection{Modules Provided}\index{modules!provided}\index{provided modules}\label{sec:buildModulesProvided}

Determination of which files provide which Fortran modules is carried out by the {\normalfont \ttfamily scripts/build/moduleDependencies.pl}\index{moduleDependencies.pl@{\normalfont \ttfamily moduleDependencies.pl}} script, which generates rules describing these dependencies (and how to build the module file by compiling the corresponding source file) and writes them to {\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Module\_Dependencies}\index{Makefile_Module_Dependencies@{\normalfont \ttfamily Makefile\_Module\_Dependencies}}.

Additionally, rules are generated which describe how to build the following classes of file:
\begin{description}
\item[{\normalfont \ttfamily *.mod.d}] Contains a list of all object files upon which the module depends. Used in constructing the final set of objects which must be linked to build an executable.
\item[{\normalfont \ttfamily *.mod.gv}] Contains a list of all source files upon which the module depends. Used in constructing \gls{graphviz} visualizations of dependencies.
\item[{\normalfont \ttfamily *.m}] Contains a list of all modules provided by the corresponding object file. Used when building object files to test whether corresponding module files have been changed---allows avoidance of recompilation cascades if modules do not need to be updated (i.e. if the modules do not differ as judged by the {\normalfont \ttfamily scripts/build/compareModules.pl}\index{findExecutables.pl@{\normalfont \ttfamily compareModules.pl}} script).
\end{description}

\subsubsection{Modules Used}\index{modules!used}\index{used modules}\label{sec:buildModulesUsed}

Determination of which files use which Fortran modules is carried out by the {\normalfont \ttfamily scripts/build/useDependencies.pl}\index{useDependencies.pl@{\normalfont \ttfamily useDependencies.pl}} script, which generates rules describing these dependencies and writes them to {\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Use\_Dependencies}\index{Makefile_Use_Dependencies@{\normalfont \ttfamily Makefile\_Use\_Dependencies}}.

Additionally, rules are generated which describe how to build the following classes of file:
\begin{description}
\item[{\normalfont \ttfamily *.d}] Contains a list of all object files upon which each object file depends. Used in constructing the final set of objects which must be linked to build an executable.
\item[{\normalfont \ttfamily *.gv}] Contains a list of all source files upon which each file depends by virtue of {\normalfont \ttfamily use} statements. Used in constructing \gls{graphviz} visualizations of dependencies.
\item[{\normalfont \ttfamily *.fl}] Contains a list of libraries upon which each object file depends. Used in constructing the final set of libraries which must be linked with the executable.
\end{description}

\subsubsection{Included Files}\index{files!include}\index{include files}\label{sec:buildIncludeDeps}

Dependencies on files due to the use of ``{\normalfont \ttfamily include}'' statements (in both Fortran and C source files) are discovered by the {\normalfont \ttfamily scripts/build/includeDependencies.pl}\index{includeDependencies.pl@{\normalfont \ttfamily includeDependencies.pl}} script, which generates rules describing these dependencies and writes them to {\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Include\_Dependencies}\index{Makefile_Include_Dependencies@{\normalfont \ttfamily Makefile\_Include\_Dependencies}}.

All Fortran and C/C++ source files in the {\normalfont \ttfamily source} directory are parsed. Files specified in any include statement in a source file are added as a dependency of that source file if unless the source file is C/C++ and the named include file can be found in either {\normalfont \ttfamily source/}, a standard system include path, or in any include path specified in the {\normalfont \ttfamily GALACTICUS\_CFLAGS}\index{environment variables!GALACTICUS_CFLAGS@{\normalfont \ttfamily GALACTICUS\_CFLAGS}}\index{GALACTICUS_CFLAGS@{\normalfont \ttfamily GALACTICUS\_CFLAGS}} or {\normalfont \ttfamily GALACTICUS\_CPPFLAGS}\index{environment variables!GALACTICUS_CPPFLAGS@{\normalfont \ttfamily GALACTICUS\_CPPFLAGS}}\index{GALACTICUS_CPPFLAGS@{\normalfont \ttfamily GALACTICUS\_CPPFLAGS}} environment variables.

\subsubsection{Parameter Dependencies}\index{parameters!dependencies}\label{sec:parameterDependencies}

All runtime parameters upon which a given executable may depend are discovered by the {\normalfont \ttfamily scripts/build/parameterDependencies.pl}\index{parameterDependencies.pl@{\normalfont \ttfamily parameterDependencies.pl}} script, which generates XML files listing these parameters and writes them to {\normalfont \ttfamily \$(BUILDPATH)/\textless executableName\textgreater.parameters.xml}\index{*.parameters.xml@{\normalfont \ttfamily *.parameters.xml}}. All Fortran and C++ files in the {\normalfont \ttfamily source} directory are parsed, and embedded parameter definitions (i.e. embedded XML blocks in lines beginning ``!\@'' for Fortran or ``//\@'' for C++ with root element {\normalfont \ttfamily inputParameter}) are identified. Any files included into a source file are also searched. In the case of Fortran source files, {\normalfont \ttfamily source/*.F90}, the corresponding preprocessed file, {\normalfont \ttfamily \$(BUILDPATH)/*.p.F90}, will be searched for parameter dependencies.

\subsection{Code Generation}

During build, code is generated automatically. Much of this is to link together functions within \glc\ (thereby permitting the modular nature of \glc), but also includes generation of various utility functions. All code generation is described below.

\subsubsection{Memory Management Functions}\label{sec:buildGenerationMemory}

Functions for allocation and deallocation of intrinsic-type arrays are generated by the {\normalfont \ttfamily memoryManagementFunctions.pl}\index{memoryManagementFunctions.pl@{\normalfont \ttfamily memoryManagementFunctions.pl}} script. Utilizing the list of such types in the {\normalfont \ttfamily \$(BUILDPATH)/allocatableArrays.xml} file (see \S\ref{sec:buildDiscoverAllocatables}) overloaded functions are created for each type. These memory management functions keep track of the total amount of memory allocated.

{\normalfont \ttfamily memoryManagementFunctions.pl}\index{memoryManagementFunctions.pl@{\normalfont \ttfamily memoryManagementFunctions.pl}}
 
\subsubsection{Unique Label Functions}

Construction of functions which provide unique labelling (see \S\ref{sec:codeUniqueLabels}) is carried out by the {\normalfont \ttfamily uniqueLabelFunctions.pl}\index{uniqueLabelFunctions.pl@{\normalfont \ttfamily uniqueLabelFunctions.pl}} script. One such function is constructed for every instance of a {\normalfont \ttfamily uniqueLabel} directive embedded in the source code. This function is constructed as follows:
\begin{enumerate}
\item Find all files on which the file with the embedded {\normalfont \ttfamily uniqueLabel} depends. (These are the ``dependent files''.)
\item Scan those files for instances of input parameters being read in.
\item Append the name of that parameter and its value (or default value) to the return value of the function on the following conditions:
\begin{enumerate}
 \item For parameters not part of a Galacticus method, always append the parameter;
 \item For parameters part of an old-style Galacticus method, append the parameter only if the specific implementation of the method to which it belongs is selected at run time;
 \item For parameters part of a new-style Galacticus method, append the parameter only if the specific implementation:
\begin{enumerate}
  \item is directly used by any dependent file;
  \item is used indirectly by virtue of being extended by some other implementation that is directly or indirectly used by any dependent file;
  \item if any dependent file uses the default implementation, and the appropriate implementation to which the parameter belongs is selected at run time.
\end{enumerate}
\end{enumerate}
\item Additionally, when appending parameters from a file, also append an MD5 digest of the file (with included files included inline, comments and leading/trailing whitespace stripped) if the appropriate option is selected in the function  call. Include in this digest any XML or HDF5 file in the data/ directory that the file accesses. Also append an MD5 digest of any other files explicitly requested in the {\normalfont \ttfamily uniqueLabel} directive.
\end{enumerate}

\subsubsection{Directives}\label{sec:codeBuildIncludeDirectives}

Generation of code to implement the functionality of {\normalfont \ttfamily include} directives is carried out by the {\normalfont \ttfamily buildCode.pl}\index{buildCode.pl@{\normalfont \ttfamily buildCode.pl}} script. This script simply reads an XML file generated by the {\normalfont \ttfamily codeDirectivesParse.pl} (see \S\ref{sec:buildDiscoveryDirectives}), calls the appropriate function to generate the necessary code, passes this through the preprocessor (see \S\ref{sec:sourceTreePreprocessor}), and outputs the result to the appropriate include file.

\subsubsection{Pre- and Post-processing}\label{sec:codeBuildPrePostProcess}

Before being compiled, all Fortran source files are passed through a tree-based preprocessor. The functionality of this preprocessor is described in \S\ref{sec:sourceTreePreprocessor}. The preprocessor is invoked on each source file by the {\normalfont \ttfamily preprocess.pl}\index{preprocess.pl@{\normalfont \ttfamily preprocess.pl}} script, which takes a {\normalfont \ttfamily source/*.F90} file as input and outputs a preprocessed file {\normalfont \ttfamily \$(BUILDPATH)/*.p.F90}.

Output from the {\normalfont \ttfamily gfortran} compiler is directed through the {\normalfont \ttfamily postprocess.pl}\index{postprocess.pl@{\normalfont \ttfamily postprocess.pl}} script. This script currently performs the following functions:
\begin{itemize}
\item Builds a map from line numbers in the preprocessed source file back to line numbers in the unpreprocessed source, and translates line numbers in any error or warnings messages from the compiler into line numbers in the unpreprocessed source;
\item Filters out incorrect warning messages about the lack of scalar finalizers (see GCC \href{https://gcc.gnu.org/bugzilla/show_bug.cgi?id=58175}{PR58175});
\item Checks for unused function attributes in the source and filters out warnings about these unused functions.
\end{itemize}

\subsection{Build Files}

All files involved in the build process are summarized below.

\begin{description}

\item[{\normalfont \ttfamily scripts/build/allocatableArrays.pl}\index{allocatableArrays.pl@{\normalfont \ttfamily allocatableArrays.pl}}:] Discovers dimensionful, allocatable, intrinsic-typed arrays in the source code, and generates a summary of their types and ranks to the file {\normalfont \ttfamily \$(BUILDPATH)/allocatableArrays.xml} (see \S\ref{sec:buildDiscoverAllocatables});
  
\item[{\normalfont \ttfamily scripts/build/buildCode.pl}\index{buildCode.pl@{\normalfont \ttfamily buildCode.pl}}:] Acts upon {\normalfont \ttfamily include} directives embedded in the source code and generates the corresponding include file (see \S\ref{sec:codeBuildIncludeDirectives});

\item[{\normalfont \ttfamily scripts/build/codeDirectivesParse.pl}\index{codeDirectivesParse.pl@{\normalfont \ttfamily codeDirectivesParse.pl}}:] Discovers and parses directives embedded in source files. Generates {\normalfont \ttfamily make} rules for the resulting dependencies, a file providing a mapping of which files contain each directive, and rule files describing how to build each include file resulting from a {\normalfont \ttfamily include} directive;

\item[{\normalfont \ttfamily scripts/build/compareModuleFiles.pl}\index{compareModuleFiles.pl@{\normalfont \ttfamily compareModuleFiles.pl}}:] Tests whether two Fortran module files are identical (ignoring timestamp information). Used to avoid updating modules when not necessary and so avoids recompilation cascades;

\item[{\normalfont \ttfamily scripts/build/includeDependencies.pl}\index{includeDependencies.pl@{\normalfont \ttfamily includeDependencies.pl}}:] Discovers (and generates rules for) dependencies between files arising from the use of ``{\normalfont \ttfamily include}'' statements (see \S\ref{sec:buildIncludeDeps});
  
\item[{\normalfont \ttfamily \normalfont \ttfamily scripts/build/executableSize.pl\index{executableSize.pl@{\normalfont \ttfamily executableSize.pl}}}:] Generates a file containing details of the size of each generated executable which can be used at runtime for memory reporting (see \S\ref{sec:buildExecutables});

\item[{\normalfont \ttfamily \normalfont \ttfamily scripts/build/findExecutables.pl\index{findExecutables.pl@{\normalfont \ttfamily findExecutables.pl}}}:] Discovers (and generates rules for) files which generate executables (see \S\ref{sec:buildExecutables});

\item[{\normalfont \ttfamily scripts/build/libraryDependencies.pl}\index{libraryDependencies.pl@{\normalfont \ttfamily libraryDependencies.pl}}:] Determines the set of libraries that must be linked with each executable, and ensures they are ordered correctly for static linking;

\item[{\normalfont \ttfamily scripts/build/memoryManagementFunctions.pl}\index{memoryManagementFunctions.pl@{\normalfont \ttfamily memoryManagementFunctions.pl}}:] Generates memory allocation/deallocation functions for allocatable intrinsic arrays (see \S\ref{sec:buildGenerationMemory});

\item[{\normalfont \ttfamily scripts/build/moduleDependencies.pl}\index{moduleDependencies.pl@{\normalfont \ttfamily moduleDependencies.pl}}:] Discovers (and generates rules for) dependencies of module files on their source file statements (see \S\ref{sec:buildModulesProvided});

\item[{\normalfont \ttfamily scripts/build/parameterDependencies.pl}\index{parameterDependencies.pl@{\normalfont \ttfamily parameterDependencies.pl}}:] Determines all parameters upon which a given executable depends (see \S\ref{sec:parameterDependencies});

\item[{\normalfont \ttfamily scripts/build/postprocess.pl}\index{postprocess.pl@{\normalfont \ttfamily postprocess.pl}}:] Postprocesses the output of the {\normalfont \ttfamily gfortran} compiler to remove spurious warnings and provide line numbers in warning/error reports that point into the unpreprocessed source files (see \S\ref{sec:codeBuildPrePostProcess});

\item[{\normalfont \ttfamily scripts/build/preprocess.pl}\index{preprocess.pl@{\normalfont \ttfamily preprocess.pl}}:] Preprocesses Fortran source code to provide various extended functionality (see \S\ref{sec:codeBuildPrePostProcess});

\item[{\normalfont \ttfamily scripts/build/uniqueLabelFunctions.pl}\index{uniqueLabelFunctions.pl@{\normalfont \ttfamily uniqueLabelFunctions.pl}}:] Builds functions which provide unique labelling of source files (see \S\ref{sec:codeUniqueLabels});

\item[{\normalfont \ttfamily scripts/build/useDependencies.pl}\index{useDependencies.pl@{\normalfont \ttfamily useDependencies.pl}}:] Discovers (and generates rules for) dependencies between files originating from module {\normalfont \ttfamily use} statements (see \S\ref{sec:buildModulesUsed});
  
\item[{\normalfont \ttfamily \$(BUILDPATH)/Makefile\_All\_Execs}\index{Makefile_All_Execs@{\normalfont \ttfamily Makefile\_All\_Execs}}:] Contains rules describing how to build the executable file for each distinct program---generated by {\normalfont \ttfamily scripts/build/findExecutables.pl}\index{findExecutables.pl@{\normalfont \ttfamily findExecutables.pl}} (see \S\ref{sec:buildExecutables});

\item[{\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Component\_Includes}\index{Makefile_Component_Includes@{\normalfont \ttfamily Makefile\_Component\_Includes}}:] Describes dependencies on the {\normalfont \ttfamily nodeComponent} class hierarchy module on include files which implement specific functionality for individual node component implementations;
  
\item[{\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Directives}\index{Makefile_Directives@{\normalfont \ttfamily Makefile\_Directives}}:] Describes rules for building include files for {\normalfont \ttfamily include} directives;

\item[{\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Include\_Dependencies}\index{Makefile_Include_Dependencies@{\normalfont \ttfamily Makefile\_Include\_Dependencies}}:] Contains rules describing the dependencies of source file on files included via ``{\normalfont \ttfamily include}'' statements---generated by {\normalfont \ttfamily scripts/build/includeDependencies.pl}\index{includeDependencies.pl@{\normalfont \ttfamily includeDependencies.pl}} (see \S\ref{sec:buildIncludeDeps});

\item[{\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Module\_Dependencies}\index{Makefile_Module_Dependencies@{\normalfont \ttfamily Makefile\_Module\_Dependencies}}:] Contains rules which describe how to build module files from their source files (see \S\ref{sec:buildModulesProvided});

\item[{\normalfont \ttfamily \$(BUILDPATH)/Makefile\_Use\_Dependencies}\index{Makefile_Use_Dependencies@{\normalfont \ttfamily Makefile\_Use\_Dependencies}}:] Contains rules which describe dependencies between source file originating from module {\normalfont \ttfamily use} statements (see \S\ref{sec:buildModulesUsed});

\item[{\normalfont \ttfamily \$(BUILDPATH)/allocatableArrays.xml}\index{allocatableArrays.xml files@{\normalfont \ttfamily allocatableArrays.xml} files}:] Contains a summary of the types and ranks of dimensionful, allocatable, intrinsic-type arrays for which memory allocation/deallocation functions will be generated (see \S\ref{sec:buildDiscoverAllocatables});

\item[{\normalfont \ttfamily \$(BUILDPATH)/directiveLocations.xml}\index{directiveLocations.xml files@{\normalfont \ttfamily directiveLocations.xml} files}:] Contains a map of which files contain each code directive (see \S\ref{sec:buildDiscoveryDirectives});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.xml}\index{*.xml files@{\normalfont \ttfamily *.xml} files}:] Contain rules for building files for {\normalfont \ttfamily include} files (see \S\ref{sec:buildDiscoveryDirectives});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.d}\index{*.d files@{\normalfont \ttfamily *.d} files}:] Contain lists of dependencies on object files, and are accumulated to find the final set of files which must be linked to build each executable (see \S\ref{sec:buildModulesProvided} \& \S\ref{sec:buildModulesUsed});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.m}\index{*.m files@{\normalfont \ttfamily *.m} files}:] Contain lists of modules provided by each object file, and are used during testing whether modules have been updated (see \S\ref{sec:buildModulesProvided});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.fl}\index{*.fl files@{\normalfont \ttfamily *.fl} files}:] Contain lists of libraries upon which each object file dependes, and are used to determine the final set of libraries which must be linked with each executable (see \S\ref{sec:buildModulesUsed});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.gv}\index{*.gv files@{\normalfont \ttfamily *.gv} files}:] Contain lists of source files upon which each file depends and are used in the construction of \gls{graphviz} representations of file dependencies (see \S\ref{sec:buildModulesProvided} \& \S\ref{sec:buildModulesUsed});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.Inc}\index{*.Inc files@{\normalfont \ttfamily *.Inc} files}:] Unpreprocessed include files generated from {\normalfont \ttfamily include} dependencies (see \S\ref{sec:buildDiscoveryDirectives});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.inc}\index{*.inc files@{\normalfont \ttfamily *.inc} files}:] Preprocessed include files generated from {\normalfont \ttfamily include} dependencies (see \S\ref{sec:buildDiscoveryDirectives});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.p.F90}\index{*.p.F90 files@{\normalfont \ttfamily *.p.F90} files}:] Preprocessed Fortran source files (see \S\ref{sec:codeBuildPrePostProcess});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.parameters.xml}\index{*.parameters.xml files@{\normalfont \ttfamily *.parameters.xml} files}:] Contain lists of parameters upon which an executable depends (see \S\ref{sec:parameterDependencies});

\item[{\normalfont \ttfamily \$(BUILDPATH)/*.size}\index{*.size files@{\normalfont \ttfamily *.size} files}:] Contain the size of each executable file built, and are used at runtime for memory reporting purposes---generated by {\normalfont \ttfamily scripts/build/executableSize.pl}\index{executableSize.pl@{\normalfont \ttfamily executableSize.pl}} (see \S\ref{sec:buildExecutables}).
\end{description}
